---
title: "KDE for Population Modelling"
author: "Michal Michalski"
date: "14/12/2020"
output: 
  html_document:
    theme: flatly
    highlight: tango
    code_download: true
    toc: true
    toc_float: 
      collapsed: true
      smooth_scroll: false
  
---
<style>
div.blue { background-color:#D3D3D3; border-radius: 5px; padding: 20px;}
</style>
<div class = "blue">
Notes on using Kernel Density Estimation for modelling population density
</div>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Packages
First, I am loading all packages necessary for the following analysis.
```{r packages, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

library(raster)# deal with raster
library(sf) # spatial class
library(mapview) # interactive map viewing
library(tmap) # cartography
library(tmaptools) # little helpers
library(dplyr) # manipulate
library(esri2sf) # get data from the server
library(spatstat) # spatial statistics
library(sp) 
library(rgdal)
library(maptools)
library(spex)
```

## Data

Data Zones Centroids
```{r}

url_dz =  "http://sedsh127.sedsh.gov.uk/arcgis/rest/services/ScotGov/StatisticalUnits/MapServer/4"

dz_cntr = esri2sf(url_dz, outFields=c("TotPop2011")) %>% st_transform(27700)

plot(dz_cntr)

```


NHS Health Boards bounadaries - above method does not work for polygons in this situation
```{r}

hb_pol = st_read("data/SG_NHS_HealthBoards_2019.shp")

plot(hb_pol)

```

## Area of Interest

Here I will select NHS Lothian Health Board and clip the Data Zones centroids to its boundaries

### Filter
```{r}

hb_lothian = hb_pol %>% 
  filter(HBName == "Lothian") %>% 
  select(-Shape_Leng, -Shape_Area)

plot(hb_lothian)

```

### Subset

```{r}

dz_lothian = dz_cntr[hb_lothian,]

plot(dz_lothian)

```
## Binning 

### GRID

```{r}
rec_grid = st_make_grid(dz_lothian, cellsize = 500, square = TRUE)

rec_grid_sf = st_sf(rec_grid)

rec_grid_sf <- rec_grid_sf %>% mutate(id = row_number())

rec_grid_join = st_join(rec_grid_sf, dz_lothian)

rec_grid_sum = rec_grid_join %>%
  group_by(id) %>%
  summarize(
    dz_count = n(),
    pop_sum  = sum(TotPop2011)
  ) %>% 
  filter(!is.na(pop_sum))
```

Map

```{r}

tmap_mode(mode = c("view"))

tm_shape(rec_grid_sum) +
  tm_fill(col = "pop_sum",
          style = "jenks",
          convert2density = TRUE) + 
  tm_borders(alpha=.8, col = "white")
```

<div class = "blue">
ESRI - Point Density (SPatial Analyst) - focal operation with movable window e.g.
1 point / (9 cells * 100m2 area per cell) = 0.0011 points per square meter (within a 3x3 search window)

https://pro.arcgis.com/en/pro-app/tool-reference/spatial-analyst/point-density.htm
</div>

## Kernel Density Estimation

### Definition

**Kernel Density Estimation (KDE)** - method of analyzing the first-order of intensity of point pattern

* the kernel
* the bandwidth
* the edge effect

https://mathisonian.github.io/kde/


### Library(spatstat)

Convert sf class to ppp class - DZ Centroids

```{r}

dz_lothian.sp <- as_Spatial(dz_lothian)

class(dz_lothian.sp)

dz_lothian.ppp <- as(dz_lothian.sp, "ppp")

class(dz_lothian.ppp)


```


Convert sf class to ppp class - HB Boundaries

```{r}

hb_lothian.sp <- as_Spatial(hb_lothian)

class(hb_lothian.sp)

hb_lothian.ppp <- as(hb_lothian.sp, "owin")

class(hb_lothian.ppp)


```

Plot the Point Pattern for Analysis

```{r}

# assign boundaries to dz ppp

Window(dz_lothian.ppp) <- hb_lothian.ppp

# plot
plot(dz_lothian.ppp, main=NULL, cols=rgb(0,0,0,.2), pch=20)

```


Create KDE

```{r}

kde_200 = density(dz_lothian.ppp, 
        sigma=200, # choose bandwith / diameter of the Kernel in the units your map is in
        eps = 100, # pixel resolution
        weights= dz_lothian.ppp$marks, # population 
        edge=TRUE, 
        varcov=NULL,
        at="pixels",
        eaveoneout=TRUE,
        adjust=1, 
        diggle=FALSE, 
        se=FALSE,
        kernel="gaussian", # choose kernel
        scalekernel=is.character(kernel), 
        positive=FALSE,
        verbose=TRUE)

kde_500 = density(dz_lothian.ppp, 
        sigma=500, # choose bandwith / diameter of the Kernel in the units your map is in
        eps = 100, # pixel resolution
        weights= dz_lothian.ppp$marks, # population
        edge=TRUE, 
        varcov=NULL,
        at="pixels",
        eaveoneout=TRUE,
        adjust=1, 
        diggle=FALSE, 
        se=FALSE,
        kernel="gaussian", # choose kernel
        scalekernel=is.character(kernel), 
        positive=FALSE,
        verbose=TRUE)

```

Plot KDE


```{r}

plot(kde_200, main = "Sigma 200m")
plot(kde_500, main = "Sigma 500m")

```

<div class = "blue">
ESRI - Kernel Density (Spatial Analyst)

https://pro.arcgis.com/en/pro-app/tool-reference/spatial-analyst/kernel-density.htm

</div>

### Extract Clusters

```{r}

kde_raster = raster(kde_500)

kde_pol = spex::polygonize(kde_raster)

kde_pol2 = kde_pol %>% 
   filter(layer > 0.001) %>% # DEFINE THRESHOLD
   summarize() 

kde_pol2 = kde_pol2 %>% st_set_crs(27700)

```

<div class = "blue">
ESRI - Raster to Polygon (Conversion)

https://pro.arcgis.com/en/pro-app/tool-reference/conversion/raster-to-polygon.htm

</div>

## Settlements

```{r}

# settlements
sett_url =  "http://sedsh127.sedsh.gov.uk/arcgis/rest/services/NRS/NRS/MapServer/9"

sett_pol = esri2sf(sett_url, outFields=c("name")) %>% st_transform(27700)

sett_lothian = sett_pol[hb_lothian,]

```

Map

```{r}

tmap_mode(mode = c("view"))

tm_shape(kde_pol2) +
  tm_fill(col = "red", alpha = 0.5) +
tm_shape(sett_lothian) +
  tm_fill(col = "blue", alpha = 0.5)
  
```



## Finally, session info

```{r}
sessionInfo()
```


&nbsp;
<hr />
<p style="text-align: center;">A work by <a href="https://github.com/topographos/">Michal Michalski</a></p>
&nbsp;
